The universal property of infinite direct sums in C$^*$-categories and W$^*$-categories

### 2019-09-23 direct_sums.tex
further minor corrections

There appear many useful universal properties for purely
algebraic structures.  Treating analytical structures
categorically is just as straightforward in some cases, but very challenging
in many other cases. It usually involves equipping the categories
under consideration themselves with certain analytical structures.
An interesting case of where this happens is the theory of
\emph{W$^*$-categories}, which model categories of Hilbert space
representations of C$^*$-algebras or W$^*$-algebras. The most
paradigmatic example of a W$^*$-category being $\Hilb$, the category
of Hilbert spaces with bounded linear maps.

@@

\begin{defn}
A complex $*$-category $(\cat{C},*)$ is called a
\begin{enumerate}
	\item \emph{normed $*$-category} if each homset~$\cat{C}(A,B)$ is a normed space such that
	    \[
		    \| ab \| \ \leq\  \|a\| \, \|b\|
	    \]
                for any~$a\colon B \to C$ and~$b\colon A \to B$.
	\item\label{Cstarcat}
    \emph{C$^*$-category} if moreover every homset is norm-complete, and the C$^*$-identity
    \[
	\|a \|^2\  =\  \|a^*a\|
     \]
	holds for every morphism $a$.
        (This turns $\cat{C}(A,A)$ into a~C$^*$-algebra.)
\item\label{Wstarcat}
    \emph{W$^*$-category} if moreover
        where every homset has a Banach space predual.

	Equivalently~\cite[Prop.~2.15]{wstarcat}, a C$^*$-category~$\cat{C}$
            is a~W$^*$-category
            if every~$\cat{C}(A,A)$ is a W$^*$-algebra
            and~$\cat{C}(A,B)$ is self dual
            as a Hilbert~$\cat{C}(A,A)$-module
            with inner product~$\langle a,b\rangle \equiv a^*b$.
\end{enumerate}
As the homset~$\cat{C}(A,A)$
    in a C$^*$-category
    is a C$^*$-algebra,
    we will freely use definitions from~C$^*$-algebras on endomorphisms,
    so we say~$a\colon A \to A$ is a \emph{projection}
        if it is a projection as an element of the~C$^*$-algebra
        $\cat{C}(A,A)$.
Several notions extend naturally to all morphisms:
    a morphism~$a \colon A \to B$ is called an \emph{isometry} if~$a^*a=1$
        (and in that case~$aa^*$ is a projection)
        and it is a \emph{unitary} if additionally~$aa^*=1$.
\end{defn}

@@

In this section, we write $\cat{C}$ for a C$^*$-category \cite{wstarcat}.
We start with a generalization of~\cite[Lemma 7]{westerbaan2016universal}
    to C$^*$-categories before discussing universal properties.

@@

\begin{lem}
Let~$a \colon A \leftrightarrows B \colon b$ be any two morphisms
    such that~$\|a \| \leq 1$, $\| b\| \leq 1$ and~$ab=1$.
    Then~$b$ is an isometry with~$a=b^*$.
\label{isometrylemma}
\end{lem}
\begin{proof}
    To start, note~$\|a ^*a\| = \|a\|^2 \leq 1$, so~$a^*a \leq 1$.
    Similarly~$b^*b \leq 1$.
    Combined~$1 = b^*a^*ab \leq b^*b \leq 1$, hence~$b^*b=1$
        and so~$b$ is an isometry.
    Similarly~$aa^*=1$.
    Now~$a (1-bb^*) a^* = aa^* - abb^*a^* = 1-1=0 \leq 0$.
    So~$a^* (1-0) a \leq bb^*$ by Lemma~\ref{contrapositionlemma}.
    Similarly~$bb^* \leq a^*a$. Hence~$bb^*=a^*a$.
    We are done: $a = aa^*a=abb^*=b^*$.
\end{proof}

@@

    The convergence $f^{(n)} \to f$ works similarly. By assumption, for every~$\eps > 0$ we have~$\left\| f^{(n)} - f^{(m)} \right\| \leq \eps$ for almost all $n,m\in\N$. In particular, we know that for any finite $S \subseteq I$,
\[
	\Bigl\| \sum_{i\in S} \bigl(f_i^{(n)} - f_i^{(m)}\bigr) \bigl(f_i^{(n)} - f_i^{(m)}\bigr)^* \Bigr\|^{1/2} \ \leq\  \eps
\]
for almost all $n,m\in\N$. Taking $m\to\infty$ gives
\[
	\Bigl\| \sum_{i\in S} \bigl(f_i^{(n)} - f_i\bigr) \bigl(f_i^{(n)} - f_i\bigr)^* \Bigr\|^{1/2} \ \leq \ \eps.
\]
Since $S$ was arbitrary, we conclude $\| f^{(n)} - f \|  \leq \eps$ for almost all $n$.
\end{proof}

The map $B \mapsto \bigoplus_i \cat{C}(A_i,B)$ is functorial in $B$ in the obvious way, yielding a functor
\[
	\bigoplus_i \cat{C}(A_i,-)  \colon \  \cat{C} \to \Ban.
\]
We now get to our main definition:

@@

@@

	We can thus try to define the \emph{direct sum with kernel $K$} to be an object representing this functor in the same way as above. For example in $\Hilb$ with $A_i = \C$ for all $i\in I$, the resulting direct sum would be the space of families $(\xi_i)_{i\in I}$ with $\xi_i\in\C$ that are $K$-square summable in that $\sum_{i,j} K_{ij} \bar{\xi}_i \xi_j < \infty$, turned into a Hilbert space with respect to the inner product with kernel $K$. We do not yet know what these direct sums with kernel will amount to in other cases.
\end{rem}

@@

We now derive \ref{biprodsum} from \ref{universalprop2}. Applying the assumption on norms with $(f_j) := (\kappa_j)$ gives $\| \sum_i \kappa_i \kappa_i^* \| = \| \id{A} \|$, and therefore $\| \kappa_j \| \leq 1$ for all $j\in I$. For fixed $j$, the family $(\delta_{ij} : A_i \to A_j)$ corresponds to some morphism $\pi_j : A \to A_j$ such that $\pi_j\circ\kappa_i = \delta_{ij}$. Again by the isometry property, we have $\| \pi_j \| = \| \id{A} \| \leq 1$. By Lemma~\ref{isometrylemma}, this implies $\pi_j = \kappa_j^*$. We therefore have $\kappa_j^*\kappa_i = \delta_{ij}$, as desired. This also implies that the $\kappa_j \kappa_j^*$ are mutually orthogonal projections. In order to also prove the completeness relation~\eqref{complete}, let $p : A\to A$ be any other projection that is orthogonal to each of the $\kappa_j \kappa_j^*$; we need to show $p = 0$. Again by naturality, under the isomorphism $\cat{C}(A,A) \cong \bigoplus_i \cat{C}(A_i,A)$, the new $p$ corresponds to the family $(p\kappa_i) = (0)$, where $p\kappa_i = 0$ follows from the assumed orthogonality. This implies indeed $p=0$ by the uniqueness part of the universal property.

>>>>

There exist many useful universal properties throughout mathematics, especially so for purely
algebraic structures.  Treating analytical structures
categorically is just as straightforward in some cases, but very challenging
in many other cases. It usually involves equipping the categories
under consideration themselves with certain analytical structures.
An interesting case of where this happens is the theory of
\emph{W$^*$-categories}, which model categories of Hilbert space
representations of C$^*$-algebras or W$^*$-algebras. The most
paradigmatic example of a W$^*$-category being $\Hilb$, the category
of Hilbert spaces with bounded linear maps.

@@

\begin{defn}
A complex $*$-category $(\cat{C},*)$ is called a
\begin{enumerate}
	\item \emph{normed $*$-category} if each homset~$\cat{C}(A,B)$ is a normed space such that
	    \[
		    \| ab \| \ \leq\  \|a\| \, \|b\|
	    \]
                for any~$a\colon B \to C$ and~$b\colon A \to B$.
	\item\label{Cstarcat}
    \emph{C$^*$-category} if moreover every homset is norm-complete, and the C$^*$-identity
    \[
	\|a \|^2\  =\  \|a^*a\|
     \]
	holds for every morphism $a$.
        (This turns $\cat{C}(A,A)$ into a~C$^*$-algebra.)
\item\label{Wstarcat}
    \emph{W$^*$-category} if moreover every homset has a Banach space predual.

	Equivalently~\cite[Prop.~2.15]{wstarcat}, a C$^*$-category~$\cat{C}$
            is a~W$^*$-category
            if every~$\cat{C}(A,A)$ is a W$^*$-algebra
            and~$\cat{C}(A,B)$ is self dual
            as a Hilbert~$\cat{C}(A,A)$-module
            with inner product~$\langle a,b\rangle \equiv a^*b$.
\end{enumerate}
As the homset~$\cat{C}(A,A)$
    in a C$^*$-category
    is a C$^*$-algebra,
    we will freely use definitions from~C$^*$-algebras on endomorphisms,
    so we say e.g.~that~$a\colon A \to A$ is a \emph{projection} if it is a projection as an element of the~C$^*$-algebra $\cat{C}(A,A)$.

Several notions extend naturally to all morphisms:
    a morphism~$a \colon A \to B$ is called an \emph{isometry} if~$a^*a=1$
        (and in that case~$aa^*$ is a projection)
        and it is a \emph{unitary} if additionally~$aa^*=1$.
\end{defn}

@@

In this section, we write $\cat{C}$ for a C$^*$-category.
We start with a generalization of~\cite[Lemma 7]{westerbaan2016universal}
    to C$^*$-categories before discussing universal properties.

@@

\begin{lem}
Let~$a \colon A \leftrightarrows B \colon b$ be any two morphisms
    such that~$\|a \| \leq 1$, $\| b\| \leq 1$ and~$ab=1$.
    Then~$b$ is an isometry with~$a=b^*$.
\label{isometrylemma}
\end{lem}
\begin{proof}
    To start, note~$\|a ^*a\| = \|a\|^2 \leq 1$, so~$a^*a \leq 1$.
    Similarly~$b^*b \leq 1$.
    Combined~$1 = b^*a^*ab \leq b^*b \leq 1$, hence~$b^*b=1$
        and so~$b$ is an isometry.
    Similarly~$aa^*=1$.
    Now~$a (1-bb^*) a^* = aa^* - abb^*a^* = 1-1=0 \leq 0$.
    So~$a^* 1 a \leq bb^*$ by Lemma~\ref{contrapositionlemma}.
    Similarly~$bb^* \leq a^*a$. Hence~$bb^*=a^*a$.
    We are done: $a = aa^*a=abb^*=b^*$.
\end{proof}

@@

    The convergence $f^{(n)} \to f$ works similarly. By assumption, for every~$\eps > 0$ we have~$\left\| f^{(n)} - f^{(m)} \right\| \leq \eps$ for sufficiently large $n,m\in\N$. In particular, we know that for any finite $S \subseteq I$,
\[
	\Bigl\| \sum_{i\in S} \bigl(f_i^{(n)} - f_i^{(m)}\bigr) \bigl(f_i^{(n)} - f_i^{(m)}\bigr)^* \Bigr\|^{1/2} \ \leq\  \eps
\]
for sufficiently large $n,m\in\N$. Taking $m\to\infty$ gives
\[
	\Bigl\| \sum_{i\in S} \bigl(f_i^{(n)} - f_i\bigr) \bigl(f_i^{(n)} - f_i\bigr)^* \Bigr\|^{1/2} \ \leq \ \eps.
\]
Since $S$ was arbitrary, we conclude $\| f^{(n)} - f \|  \leq \eps$ for sufficiently large $n$.
\end{proof}

The map $B \mapsto \bigoplus_i \cat{C}(A_i,B)$ is functorial in $B$ in the obvious way: for $g : B \to C$, a family $(f_i : A_i \to B)_{i\in I}$ gets mapped to $(gf_i : A_i \to C)_{i\in I}$, which satisfies the relevant square integrability condition because of
\[
	\sum_i (g f_i)(g f_i)^* = g \left( \sum_i f_i f_i^* \right) g^* < \infty.
\]
Thus we get a functor
\[
	\bigoplus_i \cat{C}(A_i,-)  \colon \  \cat{C} \to \Ban.
\]
We now get to our main definition:

@@

\begin{rem}
	One may wonder whether our direct sums can be obtained as \emph{weighted limits} or \emph{weighted colimits}, which are generally regarded as the adequate notion of limits and colimits in an enriched context like ours. We believe this not to be the case: using enrichment in $\Ban$, there does not seem to be any weight which would give rise to the functor $\bigoplus_i \cat{C}(A_i,-)$ from~\eqref{AiB}. The main reason is that mere $\Ban$-enrichment does not have access to the involution, which however is a crucial piece of structure in~\eqref{AiB}.
\end{rem}

@@

	We can thus define the \emph{direct sum with kernel $K$} to be an object representing this functor in the same way as above. For example in $\Hilb$ with $A_i = \C$ for all $i\in I$, the resulting direct sum would be the space of families $(\xi_i)_{i\in I}$ with $\xi_i\in\C$ that are $K$-square summable in that $\sum_{i,j} K_{ij} \bar{\xi}_i \xi_j < \infty$, turned into a Hilbert space with respect to the inner product with kernel $K$. We do not yet know what these direct sums with kernel will amount to in other cases.
\end{rem}

@@

We now derive \ref{biprodsum} from \ref{universalprop2}. Applying the assumption on norms with $(f_j) := (\kappa_j)$ gives $\| \sum_i \kappa_i \kappa_i^* \| = \| \id{A} \|$, and therefore $\| \kappa_j \| \leq 1$ for all $j\in I$. For fixed $j$, the family $(\delta_{ij} : A_i \to A_j)_{i\in I}$ corresponds to some morphism $\pi_j : A \to A_j$ such that $\pi_j\circ\kappa_i = \delta_{ij}$. Again by the isometry property, we have $\| \pi_j \|^2 = \| \sum_i \delta_{ij} \delta_{ij}^* \| = \| \id{A_j} \| \leq 1$. By Lemma~\ref{isometrylemma}, this implies $\pi_j = \kappa_j^*$. We therefore have $\kappa_j^*\kappa_i = \delta_{ij}$, as desired. This also implies that the $\kappa_j \kappa_j^*$ are mutually orthogonal projections. In order to also prove the completeness relation~\eqref{complete}, let $p : A\to A$ be any other projection that is orthogonal to each of the $\kappa_j \kappa_j^*$; we need to show $p = 0$. Again by naturality, under the isomorphism $\cat{C}(A,A) \cong \bigoplus_i \cat{C}(A_i,A)$, the new $p$ corresponds to the family $(p\kappa_i) = (0)$, where $p\kappa_i = 0$ follows from the assumed orthogonality. This implies indeed $p=0$ by the uniqueness part of the universal property.
